Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-10 article.Rmd
Removed outdated stuff

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged “a sea change in the [historical] profession with respect to gender” before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck’s report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession. What follows is a quantitative supplement to Lunbeck’s analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period.^[http://www.historians.org/Documents/About%20AHA%20and%20Membership/CWH-Report_5.20.05.pdf]

Our analysis focuses on one of the bedrocks of the historical profession: scholarly research. We begin with the history dissertation, often the defining scholarly output of a historian’s early career. The completion of a dissertation marks a standardized moment of transition out of the “training” phase of the historical profession. Data supplied by ProQuest provided roughly [80,000] PhD(?) dissertations completed since [1950]. Identifying the gender of their authors gives a sense for the number of women and men who are completing PhD-level training in history each year. Our program uses what year a historian wrote their dissertation to estimate a period for when they might have been born: a historian who completed their dissertation in 1980 was likely born between ____ and _____. Using this temporal information, we are better able to infer their gender and chart how the larger representation of women and men changes year-by-year.

@@

Future work:

Both the Python script and the R package are easily extensible to allow additional data sets, say from other countries. (Are there other data sets, say for Great Britain, http://data.gov.uk/data/search?q=names ??) And eventually perhaps US census data, which would solve the 19th century problem.

>>>>

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck’s report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period.^[http://www.historians.org/Documents/About%20AHA%20and%20Membership/CWH-Report_5.20.05.pdf]

Our analysis focuses on one of the bedrocks of the historical profession: scholarly research. We begin with the history dissertation, often the defining scholarly output of a historian's early career. The completion of a dissertation marks a standardized moment of transition out of the "training" phase of the historical profession. Data supplied by ProQuest provided roughly [80,000] PhD(?) dissertations completed since [1950]. Identifying the gender of their authors gives a sense for the number of women and men who are completing PhD-level training in history each year. Our program uses what year a historian wrote their dissertation to estimate a period for when they might have been born: a historian who completed their dissertation in 1980 was likely born between ____ and _____. Using this temporal information, we are better able to infer their gender and chart how the larger representation of women and men changes year-by-year.

@@
